TrenchR: an R package for transparent environmental and ecological biophysics

### 2021-04-30 ManuscriptTe.Rmd
Update ms

---
title: "Manuscript Te example"
author: "Lauren Buckley"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
  rmarkdown::html_vignette: default
---

@@

```{r}
##Load TrenchR package
library(TrenchR)
library(maptools)
library(ggplot2)
library(reshape)
```

#APPENDIX S1
We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR. We first use a times series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness. We then use a time series of air and surface temperature, windspeed, and solar radiation collected a a single height to implement and test a biophysical model for grasshoppers.

#Microclimate models
```{r}
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
C1all=read.csv("microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
```

We first examine temperature and wind speed profiles and use them to estimate surface roughness.
```{r}
#ANALYZE SURFACE ROUGHNESS AND PROFILES
C1all$day=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$mday
C1all$hour=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$hour

@@

We can use wind speed variation across elevations to estimate surface roughness across hours.
```{r}
#surface roughness
z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.0,1.5))

@@

We calculate the temperature and wind speed profiles for a given hour using multiple functions in TrenchR.
```{r}
z.seq <- seq(0, 2, by = 0.1)

#subset to day and hour
C1.sub= subset(C1.dayhr, C1.dayhr$day==10)
hr=10

#estimate profiles
#temperature
t.seq1= sapply(z.seq, FUN=air_temp_profile_neutral, T_r=C1.sub[hr,"Air0.5m"], z0=z0[hr], zr=0.5, T_s=C1.sub[hr,"Soil"])

@@

#windspeed
u.seq1= sapply(z.seq, FUN=wind_speed_profile_neutral, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])

@@

u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))

#replace infinity with N
t.seq1[which(!is.finite(t.seq1))]=NA
```

We then plot the temperature and wind speed profiles.
```{r}
plot(t.seq1, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
points(t.seq2, z.seq, type = "l", lty="dashed")
points(t.seq3, z.seq, type = "l", lty="dotted")
points(C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")

@@

#Biophysical models
We first read and process time series of environmental data.
```{r}
#Grasshopper data
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
spec.dat=read.csv("SpecData.csv")

@@

#Calculate zenith
dat$psi= zenith_angle(dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)

@@

@@

#scale temperature to ground
dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)

#cut nas
dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass")]
dat= na.omit(dat)
```

We then use biophysical models to estimate body temperatures based on air and surface temperatures, wind speed, and solar radiation. We first use a biophysical model specifically configured for grasshoppers.
```{r}
##Calculate Te for each species
dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=0.0211, Acondfact=0.01)
#fix infinite
dat$Te[is.infinite(dat$Te)]<-NA
#ditch big numbers for now
dat$Te[dat$Te>100]<-NA
```

We then implement a generic energy budget. We first illustrate the estimation of convective heat transfer coefficients as a function of wind speed.
```{r}
#Estimate heat transfer coefficient using empirical relationship
dat$H_L=heat_transfer_coefficient(V=dat$Wind,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")

#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
H_L1= heat_transfer_coefficient_approximation(V=dat$Wind,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
H_L2= heat_transfer_coefficient_simple(V=dat$Wind,D=0.0211)

#surface area
A= sa_from_length(l=0.0211)
```

We then implement the energy budget as follows.
```{r}
#Gates calculation
TeGates= rep(NA, nrow(dat) )

for(ind in 1:nrow(dat) ){
  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.66, psa_ref=1-0.66, psa_air=1, psa_g=0.01, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
  if(is.numeric(out))TeGates[ind]=out
}
dat$TeGates=TeGates-273.15
```

We aggregate the data and plot average conditions across sites.
```{r}
#aggregate by hour
dat.h= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour), FUN=mean, na.rm=TRUE))
#aggregate by day and hour
dat.dh= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour, dat$J), FUN=mean, na.rm=TRUE))

@@

#plot
dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )

ggplot(data=dat.s, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~Elev)

@@

```{r}
#plot by date time
dat.dh$Jhr= dat.dh$J +dat.dh$hour/24

@@

```{r}
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/figures/")

>>>>

---
title: "Appendix S1"
author: "Buckley et al."
output:
  html_document: default
  pdf_document: default
  word_document: default
  rmarkdown::html_vignette: default
---

@@

```{r, include=FALSE}
##Load TrenchR package
library(TrenchR)
library(maptools)
library(ggplot2)
library(reshape)
```

# APPENDIX S1
We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR. We first use a times series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness. We then use a time series of air and surface temperature, windspeed, and solar radiation collected at a single height to implement and test a biophysical model for grasshoppers.

# Microclimate models
```{r, include=FALSE}
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
C1all=read.csv("microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
```

```{r, include=FALSE}
#ANALYZE SURFACE ROUGHNESS AND PROFILES
C1all$day=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$mday
C1all$hour=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$hour

@@

We first examine how temperature and windspeed profiles can be used to scale conditions to the heights of organisms. We first use wind speed variation across hours and heights to estimate surface roughness.
```{r}
#estimate surface roughness based on wind velocity (m/s) at a vector of reference heights, zr (m)
z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.0,1.5))

@@

We then calculate the temperature and wind speed profiles for a given hour using multiple functions in TrenchR.
```{r}
#subset to day and hour
C1.sub= subset(C1.dayhr, C1.dayhr$day==10)
hr=10

#vector of heights
z.seq <- seq(0, 2, by = 0.1)

#estimate profiles using three different functions including the assumption of neutral and segmented profiles. 
#T_r	and u_r are temperatures (C) and windspeeds (m/s), respectively, at reference height zr (m).
#z0	is surface roughness (m); z is the height to scale to (m).
#T_s	is surface temperature (C).

#temperature profiles
t.seq1= sapply(z.seq, FUN=air_temp_profile_neutral, T_r=C1.sub[hr,"Air0.5m"], z0=z0[hr], zr=0.5, T_s=C1.sub[hr,"Soil"])

@@

#windspeed profiles
u.seq1= sapply(z.seq, FUN=wind_speed_profile_neutral, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])

@@

u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))
```

The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground. Profiles can be fit as a single smooth function or segments.
```{r, echo=FALSE}
#replace infinity with N
t.seq1[which(!is.finite(t.seq1))]=NA

plot(t.seq1, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
points(t.seq2, z.seq, type = "l", lty="dashed")
points(t.seq3, z.seq, type = "l", lty="dotted")
points(C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")

@@

# Biophysical models
```{r, include=FALSE}
#Grasshopper data
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
spec.dat=read.csv("SpecData.csv")

@@

@@

We next illustrate estimating operative environmental temperatures, the estimated body temperatures for grasshoppers in the specified environments. We first read and process time series of environmental data. We estimate the zenith angle, the angle between the sun's rays and the vertical, across Julian days and hours.
```{r}
#Calculate zenith
dat$psi= zenith_angle(dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)

@@

#scale temperature T_r from reference height zr=0.5m to grasshopper height based on windspeed u_r, surface temperature T_s, and surface roughness z0.
dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)
```

```{r, include=FALSE}
#cut nas
dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass")]
dat= na.omit(dat)
```

We then use biophysical models to estimate body temperatures based on air and surface temperatures, wind speed, and solar radiation. We first use a biophysical model specifically configured for grasshoppers.
```{r, warning=FALSE}
##Calculate operative temperature, Te
dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=0.0211, Acondfact=0.01)

#T_a	and T_g are air and surface temperatures (C), respectively; u	is wind speed (m/s); 
#H is total (direct + diffuse) solar radiation flux (W/m^2); 
#K_t	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere;
#psi is solar zenith angle (degrees); L	is grasshopper length (m);
#Acondfact is the proportion of the grasshopper surface area in contact with the ground.
```

```{r, include=FALSE}
#fix infinite
dat$Te[is.infinite(dat$Te)]<-NA
#ditch big numbers for now
dat$Te[dat$Te>100]<-NA
```

We then implement a generic energy budget. We first illustrate the estimation of convective heat transfer coefficients as a function of wind speed.
```{r}
#Estimate heat transfer coefficient using empirical relationship
dat$H_L=heat_transfer_coefficient(V=dat$Wind,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")
#where V is	air velocity (m/s); D is the characteristic dimension (m);
#K is the thermal conductivity of air (W m^-1 K^-1);
#nu is the kinematic viscosity of air (m^2 s^-1)

#Also estimate using a spherical approximation and a simplified version of the approximation.
H_L1= heat_transfer_coefficient_approximation(V=dat$Wind,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
H_L2= heat_transfer_coefficient_simple(V=dat$Wind,D=0.0211)

#surface area from length(m)
A= sa_from_length(l=0.0211)
```

We then implement the energy budget as follows.
```{r, results = FALSE}
TeGates= rep(NA, nrow(dat) )

for(ind in 1:nrow(dat) ){
  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.66, psa_ref=1-0.66, psa_air=0.99, psa_g=0.01, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
  if(is.numeric(out))TeGates[ind]=out
}
#where psa_dir, psa_ref, psa_air, psa_g are the proportions surface area exposed to direct radiation from the sky (or enclosure), reflected radiation from the ground, and air, and in contact with the ground, respectively;
#Qabs is the solar and thermal radiation absorbed (W);
#epsilon is longwave infrared emissivity of skin (proportion);
#H_L is the convective heat transfer coefficient (W m^-2 K^-1);
#ef	is the enhancement factor, used to adjust H_L to field conditions;
#K is thermal conductivity. 
```

We aggregate the data and plot average conditions across sites.
```{r, echo=FALSE, warning=FALSE}
dat$TeGates=TeGates-273.15

#aggregate by hour
dat.h= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour), FUN=mean, na.rm=TRUE))
#aggregate by day and hour
dat.dh= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour, dat$J), FUN=mean, na.rm=TRUE))

@@

#plot
dat.pl= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
dat.pl$variable= as.character(dat.pl$variable)

#change names
dat.pl[which(dat.pl$variable=="Temp"),"variable"]<-"Ta"
dat.pl[which(dat.pl$variable=="SoilTemp"),"variable"]<-"Ts"
dat.pl[which(dat.pl$variable=="SanguinipesTemp"),"variable"]<-"Te observed"
dat.pl[which(dat.pl$variable=="Te"),"variable"]<-"Te"
dat.pl[which(dat.pl$variable=="Tgrass"),"variable"]<-"Ta grasshopper height"
dat.pl[which(dat.pl$variable=="TeGates"),"variable"]<-"Te Gates"

#make factor
dat.pl$variable= factor(dat.pl$variable, levels=c("Ta","Ts","Ta grasshopper height","Te","Te Gates","Te observed"))
#make label
dat.pl$elev.lab= paste(dat.pl$Elev,"m", sep=" ")
#make Ta, Te classification
dat.pl$Ttype="input"
dat.pl$Ttype[which(dat.pl$variable %in% c("Te","Te Gates","Te observed") )]="estimates"

ggplot(data=dat.pl, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~elev.lab)+theme_classic()+ theme(legend.position="bottom")

@@

```{r, include=FALSE}
#plot by date time
dat.dh$Jhr= dat.dh$J +dat.dh$hour/24

@@

```{r, include=FALSE}
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/figures/")

### 2021-04-30 Manuscript_full.Rmd
Update ms

1.  Much understanding of climate change responses relies on the
    assumption that organismal body temperatures are equal to temporally
    averaged air temperatures high above the ground. However, most
    organisms live near the ground and acute exposure to solar radiation
    can substantially elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses
    (<https://github.com/trenchproject/TrenchR>). The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation.
3.  [Main results]
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting and includes tutorials and well as a series of
    educational modules.

@@

Many responses of organisms and ecosystems to climate change are
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperate at weather station height (usually 2m). However, air
temperatures near the ground where most organisms reside can be
considerably warmer and absorption of solar radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
@kearney2009]. Acute thermal stress events are often more relevant to
the physiology and energy balance of organisms than daily, monthly, or
annually aggregated environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package as a component of the TrEnCh project
(<https://trenchproject.github.io>), which builds computational and
visualization tools to Translate Environmental Change into organismal
responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of the
tools has not kept pace with research on climate change responses.
Adoptions of biophysical approaches are expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr]. Increasing dissemination of computational tools
is facilitating adoption. The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020] models. Many of
the approaches are similar between the packages, but TrenchR focuses on
simple functions to aid the understanding and transparency of
biophysical approaches. The TrenchR functions allow for modular use and
easy adaptation to a variety of organismal and research questions. Some
energy budget calculation components of biophysical models are also
available in the ThermImage package [@tattersall2017]. Although many of
the TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We have minimized inclusion of plant
specific biophysical models since they are the focus of several R
packages [tealeaves: [@muir2019tea]; plantecophys:
[@duursma2015plantecophys]].

@@

Functions were adapted from biophysical ecology textbooks
[@campbell2000introduction; @gates1980biophysical] and research
articles. The source reference is included in each function's header. We
have incorporated default parameters (e.g., organism emissivity or heat
conduction rates) when general values are available. Many functions
additionally include comments that describe potential parameterizations.
The biophysical ecology texts [@campbell2000introduction;
@gates1980biophysical] include many tables with parameter values and we
provide several tables in TrenchR [Table 8.2XXXX].

@@

We introduce the function in several vignettes. A good place to start is
the *Allometry and conversions* tutorial, which provides tools for
preparing data such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* tutorial
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and windspeed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we
have updated a series of educational modules entitled "Physical
processes in ecosystems". We intend the modules to provide a more
contained and accessible introduction to microclimate and biophysical
modelling than that included in classic textbooks
[@campbell2000introduction; @gates1980biophysical]. The modules provide
less detailed coverage of many of the topics contained in the books. The
series of 14 modules starts with an overview of the calculus and physics
principles underlying the modelling. Thermodynamics and energy budget
modelling are then introduced with detailed examples. Modules address
the climate space concept, operative temperatures, biophysical models
for leaves and sheep, and heat flow in soils. We have revised the
modules (which originated from an NSF training grant in 1979 and lacked
broad distribution) to include R code and utilization of TrenchR
functions. We expanded the original series of tutorials to include
introductions by Michael Kearney to the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
modules are provided in html form and available for download as a pdf at
the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
R markdown files are available in Github at
<https://github.com/trenchproject/TrenchRmodules>.

## Examples

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
incorporated in the *Using energy balances to estimate body
temperatures* tutorial. We first estimate the energy budget with
component functions and we then use an integrated biophysical model to
estimate operative environmental temperatures. We start by generating
environmental inputs. See Appendix S1 for another example examining a
time series of microclimate data and applying and testing the
biophysical models.

We will use the energy budget to estimate the lizard's body temperature,
$T_b$, which can depart dramatically from the air temperatures due to
energy exchange with the environment. Heat is gained from absorbing
solar and thermal radiation and from metabolic reactions. Heat is lost
through the organism's emission of radiation and the evaporation of
water. The organism exchanges heat with the surrounding air or water via
convection and with the ground via conduction. The balance of these heat
exchanges can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement].

Let's assume the lizard is in an unshaded location where a weather
station at standard height (2m) indicates that the daily air temperature
varies from a minimum of 10°C to a maximum of 25°C and the wind speed
averages 1 m/s. The soil surface temperature varies from a minimum of
15°C to a maximum of 30°C. We assume that atmospheric transmissivity
$\tau=0.7$ and albedo $\rho=0.6$

@@

Let us first prepare the environmental data for analysis. We will need
to estimate hourly air and soil temperatures and radiation. We start by
estimating the day of year and the timing of sunrise and sunset:

@@

doy= day_of_year("2018-06-01", format= "%Y-%m-%d") 
snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
dayl= daylength(lat=lat, doy=doy) #estimate daylength
tr= snoon-dayl/2 #time of sunrise
ts= snoon+dayl/2 #time of sunset
```

@@

### Plot

We next plot the diurnal trends in environmental data and predicted body
temperatures.

@@

*Figure 1*. Predicted lizard body temperatures (Te) as a function of air
temperatures (Ta) at 2m and lizard height (0.02m), surface (Ts)
temperatures (°C), and solar radiation ($W/m^2$). We depict body
temperatures using two general energy budgets [solid: Tb_Gates();
dotted: Tb_CampbellNorman()] and a lizard specific biophysical model
[dashed: Tb_lizard()].

@@

TrenchR is intended to promote understanding of how organisms interact
with their environment and the consequences for physiology, energetics,
behavior, and demography. Our example implementation highlights the
importance of considering organismal body temperatures, rather than air
temperatures, when examining thermal stress and other responses to
environmental variability and change. TrenchR currently focuses on heat
balances but may be expanded to include water balances. Simple functions
can be combined as needed to produce comprehensive and transparent
models for biophysical ecology and evolution. The resultant models are
likely to be sufficiently detailed for many applications, but users are
referred to NicheMapR for more detailed biophysical models
[@kearney2017nichemapr; @kearney2020nichemapr]. Making classic
biophysical ecology techniques more accessible will allow researchers to
take advantage of rapidly accumulating data on environmental conditions
and organismal traits to understand and predict ecological and
evolutionary responses to variable and changing environments.

>>>>

1.  Much understanding of responses to climate change and variability relies on the
    assumption that organismal body temperatures are equal to temporally
    averaged air temperatures high above the ground. However, most
    organisms live near the ground and acute exposure to solar radiation and thermal extremes
    can substantially elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the organismal and environonmental parameters needed for biophysical ecology.
3.  TrenchR focuses on relatively simple and modular functions so users can create transparent and flexible biophysical models.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. To further this aim, we offer a series of educational modules to introduce the field of biophysical ecology.

@@

Many responses of organisms and ecosystems to climate change are
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperate at weather station height (usually 2m). However, air
temperatures near the ground where most organisms reside can be
considerably warmer and absorption of solar radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
@kearney2009]. Applications are increasingly demonstrating the importance of accounting for microclimate variation [@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].  Acute thermal stress events are often more relevant to
the physiology and energy balance of organisms than daily, monthly, or
annually aggregated environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package (<https://github.com/trenchproject/TrenchR>) as a component of the TrEnCh project
(<https:www.trenchproject.com>), which builds computational and
visualization tools to Translate Environmental Change into organismal
responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of the
tools has not kept pace with research on climate change responses.
Adoptions of biophysical approaches are expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr]. Increasing dissemination of computational tools
is facilitating adoption. The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models. Many of
the approaches are similar between the packages, but TrenchR focuses on
simple functions to aid the understanding and transparency of
biophysical approaches. The TrenchR functions allow for modular use and
easy adaptation to a variety of organisms and research questions. Some
energy budget calculation components of biophysical models are also
available in the ThermImage package [@tattersall2017]. Although many of
the TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We have limited inclusion of plant
specific biophysical models since they are the focus of several R
packages [tealeaves: [@muir2019tea]; plantecophys:
[@duursma2015plantecophys]].

@@

Functions were adapted from biophysical ecology textbooks
[@campbell2000introduction; @gates1980biophysical] and research
articles. The source reference is included in each function's header. We
have incorporated default parameters (e.g., organism emissivity or heat
conduction rates) when general values are available. Many functions
additionally include comments that describe potential parameterizations.
The biophysical ecology texts [@campbell2000introduction;
@gates1980biophysical] include many tables with parameter values and we
provide several data tables in TrenchR.

@@

We introduce the functions in several vignettes. A good place to start is
the *Allometry and conversions* vignette, which provides tools for
preparing data such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* vignette
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and windspeed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we
have updated a series of educational modules entitled *Physical
Processes in Ecosystems*. We intend the modules to provide a more
contained and accessible introduction to microclimate and biophysical
modelling than that included in classic textbooks
[@campbell2000introduction; @gates1980biophysical]. The modules provide
less detailed coverage of many of the topics contained in the books. The
series of 14 modules starts with an overview of the calculus and physics
principles underlying the modelling. Thermodynamics and energy budget
modelling are then introduced with detailed examples. Modules address
the climate space concept, operative temperatures, biophysical models
for leaves and sheep, and heat flow in soils. We have revised the
modules (which originated from an NSF training grant in 1979 and lacked
broad distribution) to include R code and utilization of TrenchR
functions. We expanded the original series of tutorials to include
introductions by Michael Kearney to the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
modules are provided in html form and available for download as a pdf at
the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
R markdown files are available in Github at
<https://github.com/trenchproject/TrenchRmodules>.

## Implementation

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2020, in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
incorporated in the *Using energy balances to estimate body
temperatures* tutorial. We first estimate the energy budget with
component functions and we then use an integrated biophysical model to
estimate operative environmental temperatures. We start by generating
environmental inputs. See Appendix S1 for another example examining a
time series of microclimate data and applying and testing the
biophysical models.

We will use the energy budget to estimate body temperature,
$T_b$, which can depart dramatically from the air temperatures due to
energy exchange with the environment. Heat is gained from absorbing
solar and thermal radiation and from metabolic reactions. Heat is lost
through the organism's emission of radiation and the evaporation of
water. The organism exchanges heat with the surrounding air or water via
convection and with the ground via conduction. The balance of these heat
exchanges can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement].

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2m) indicates that the daily air temperature
varies from a minimum of 10°C to a maximum of 25°C and the wind speed
averages 1 m/s. The soil surface temperature varies from a minimum of
15°C to a maximum of 30°C. We assume that atmospheric transmissivity
$\tau=0.7$ and albedo $\rho=0.6$

@@

We first prepare the environmental data for analysis. We will need
to estimate hourly air and soil temperatures and radiation. We start by
estimating the day of year and the timing of sunrise and sunset:

@@

doy= day_of_year("2020-06-01", format= "%Y-%m-%d") 
snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
dayl= daylength(lat=lat, doy=doy) #estimate daylength
tr= snoon-dayl/2 #time of sunrise
ts= snoon+dayl/2 #time of sunset
```

@@

@@

* **Figure 1**. Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation. Air temperatures (Ta, °C) at lizard height (0.02m) are predicted to exceed air temperatures at 2m and to be similar to surface temperatures (Ts). We estimate body temperatures using two general energy budgets [solid: Tb_Gates();
dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.*

@@

TrenchR is intended to promote understanding of how organisms interact
with their environment and consequences for physiology, energetics,
behavior, and demography. Our example implementation highlights the
importance of considering organismal body temperatures, rather than air
temperatures, when examining thermal stress and other responses to
environmental variability and change. TrenchR currently focuses on heat
balances but may be expanded to include water balances. Simple functions
can be combined as needed to produce comprehensive and transparent
models for biophysical ecology and evolution. The resultant models are
likely to be sufficiently detailed for many applications, but users are
referred to NicheMapR for more detailed biophysical models
[@kearney2017nichemapr; @kearney2020nichemapr]. We focus on models that predict equilibrium conditions for simplicity. Such models generally do not present computational challenges so our models are not optimized for computational efficiency. Making classic
biophysical ecology techniques more accessible will allow researchers to
take advantage of rapidly accumulating data on environmental conditions
and organismal traits to understand and predict ecological and
evolutionary responses. Considering how organisms experience their environment is central to understanding responses to variable and changing environments [@suggitt2018extinction].
